---
title: "intro to r project"
author: "jodi"
date: "2024-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1 Cleaning

In this section, we aim to check for missing and duplicate values and remove columns that are insignificant in our univariate, bivariate and multivariate analysis.

We start off by loading libraries and importing the dataset we will be using for this analysis.

```{r}
library(dplyr)
library(tidyr)
Census_Income <- read.csv("adult.csv")
```

```{r}
# Replace '?' elements with NA
Census_Income <- data.frame(lapply(Census_Income, function(x) ifelse(x == '?', NA, x)))

# Drop columns native.country, capital.gain, and capital.loss
Census_Income <- subset(Census_Income, select = -c(native.country, capital.gain, capital.loss))

# Remove missing values from the column occupation
Census_Income <- Census_Income[complete.cases(Census_Income$occupation), ]
```

## 2 Basic visualisation

To decide on which variables we will focus on in our analysis, we will run some basic statistical tests to understand the variables better.

```{r}

```

## 3 Univariate analysis

You can also embed plots, for example:

```{r}

```

## 4 Multivariate analysis
You can also embed plots, for example:

```{r}

```

## 5 Graphical representations

You can also embed plots, for example:

```{r}

```
